---
title: "Clustering multivariate tail dependence with CeCl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

TODO (marginal)
- Explain that currently only supported for single predictor (location, individual, etc) (also add to package!!)
- Fix plotting methods for evgam fits! Only working for ismev
- Allow different `qgam` and `evgam` formulae for each variable
- Is the `quantile` thresholding method calculated per location or overall for a single variable?
- Remove titles for marginal plots
- Add plotting functions for transformed marginal data
- For user-inputted marginals, fix printing variables
- Add `as_cecl_marg` method that also takes in GPD parameters and performs transformation
- For marginal diagnostic plots, standardise x-axis/y-axis limits
  - Also add uncertainty bounds for PP/QQ plots, and allow log scale for return level plots
  
TODO (dependence)
- Implement bootstrap
- Allow different starting values for each location
- Only return n rows of coefs for summary of fits
- Allow back transformation from standard Laplace to original scale for quantile plots
- `plot_scatter` for ggplot should have alpha and beta, not a and b as axis labels
- Add a `predict` method

TODO (distance)
- For image plot, add option to suppress axis labels (x and y separately)
- For ggplot image, have same colour scale as clustering method uses
- Add `lambda` parameter to weigh variables differently in distance calculation

TODO (clustering)
- For image plot, coloured image plot has y-axis labels too close

This vignette shows how to use the `CeCl` package to perform clustering for 
multivariate extremes based on the conditional extremes (CE) model of Heffernan and Tawn (2004) [-@Heffernan2004]. 
We begin with performing the marginal transformation using the 
`cecl_marg` function, which has several in-built options for marginal fitting.
Optionally, the user can also provide their own transformed data using `as_cecl_marg`. 
Next, we fit the CE model using the `cecl_dep` function. 
Similarly, the user can also provide their own CE parameter estimates
using `as_cecl_dep`.
<!-- This function borrows much of it's functionality from the `texmex` package [@texmex]. -->
Finally, we perform clustering of the dependence structures using the `cecl_cluster` function.

Note that this package is still in the early stages of development, and may 
have some bugs or incomplete features and undergo significant changes in the future;
please report any issues or suggestions on the [GitHub issues page](https:/github.com/potoole7/CeCl/issues).

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(CeCl) # main package
library(copula) # for generating data
library(ggplot2) 
library(patchwork) # for combining ggplots
library(dplyr)
```

## Simulate Data 

To mimic a real live scenario, we simulate some data consisting of three variables observed for 12 "locations", clustered into 3 groups of 4 sites each.
Note that although we use spatial terminology here, the method is applicable to any multivariate data, for example, financial assets, or experimental units in a clinical trial.

We simulate multivariate data using t-copulas with different correlation parameters, for generalised Pareto (GP) marginals.
The correlation parameter for each group/cluster is set to be different, so that we can see if our clustering method is able to recover these groups based on their tail dependence structures.
```{r, simulate_data}
# metadata
n_vars <- 3 # number of variables
n_locs <- 12 # number of locations
n_clust <- 3 # number of clusters
n <- 1000 * (12 / 3) # total number of observations per group
df_t <- 3 # degrees of freedom for t-copula
cor_t_vec <- c(0.1, 0.5, 0.9) # correlation parameters for each group

# Function to generate multivariate t data with specified correlation
gen_t <- \(cor_t) {
  # generate data
  cop_t <- tCopula(param = cor_t, dim = n_vars, df = df_t, dispstr = "ex")
  u <- rCopula(n, cop_t)
  data.frame(apply(u, 2, qgpd, xi = -0.05, sigma = 1, u = 0))
}

# Generate data for two different correlation settings
set.seed(123)
data <- do.call(rbind, lapply(cor_t_vec, gen_t))
# Add dummy location names
# data$name <- rep(paste0("loc_", 1:10), each = 200)
data$name <- rep(paste0("loc_", 1:12), each = nrow(data) / n_locs)

# Visualise the data
data |>
  mutate(clust = case_when(
    name %in% paste0("loc_", 1:4) ~ "A",
    name %in% paste0("loc_", 5:8) ~ "B",
    TRUE ~ "C"
  )) |>
  ggplot(aes(x = X1, y = X2, colour = clust)) +
  facet_wrap(~ clust) +
  geom_density_2d(show.legend = FALSE) +
  cecl_theme()
```

Visualising the data, we can see that the three groups have different dependence structures, with group A having weak dependence, group B having moderate dependence, and group C having strong dependence.

...
More complex simulation examples using mixtures of Gaussian and t-copulas can
be found in [@otoole2025].

## Marginal Transformation

Generally, the CE model is fitted to data transformed to standard Laplace margins.
To perform this transformation, the marginal distributions are first estimated. 
Generally, this is done by fitting a Generalized Pareto Distribution (GPD) to the upper tail of each variable, and using the empirical CDF for the bulk of the data.

Another important aspect of marginal fitting is threshold selection for the GPD fit.
This is a crucial and difficult step in any extreme value analysis, and we do not attempt to provide a comprehensive solution here.
We recommend starting with the functions in `ismev` or `texmex` for graphical threshold selection, using mean residual life plots and threshold stability plots.
turning to the wealth of other more advanced methods available in the literature. 
Once a suitable threshold quantile or value has been selected, this can be passed to the `cecl_marg` function using the `thresh_method` and `thresh_args` arguments, which supports:
- Value thresholding (`"value"`), where the user specifies a fixed threshold value for each variable,
- Quantile thresholding (`"quantile"`), where the user specifies a fixed quantile for each variable,
- Regression thresholding using quantile GAMs (`"qgam"`), where the user specifies a formula and quantile for fitting a quantile GAM using the `qgam` package [-@qgam]. Note that this method allows for non-stationary thresholds that vary with covariates. 

Here, we demonstrate how to use each of these thresholding methods with the `cecl_marg` function:
```{r, marginal_thresh}
# 1: Specific value
thresh_val <- cecl_marg(
  data,
  thresh_method = "value",
  thresh_args = apply(data[, c("X1", "X2", "X3")], 2, quantile, probs = 0.9),
  thresh_only = TRUE
)

# 2. quantile
thresh_q <- cecl_marg(
  data,
  thresh_method = "quantile",
  thresh_args = 0.9, 
  thresh_only = TRUE
)

# 3. Regression
# TODO Fix!
# thresh_reg <- cecl_marg(
#   data,
#   thresh_method = "qgam",
#   thresh_args = list(
#     f      = "response ~ name", 
#     qu     = .9
#   ),
#   thresh_only = TRUE,
#   ret_obj = TRUE # return `qgam` object
# )

class(thresh_q)
```

By specifying `thresh_only = TRUE`, the `cecl_marg` function will only perform thresholding and return an object of class `cecl_thresh`, which contains the threshold values and the thresholded data, and, in the case of regression thresholding, the fitted `qgam` objects.
See `?cecl_marg` for more details on the outputs and available argument options.
Currently, no methods are implemented for the `cecl_thresh` class, but this may be added in future versions.

Next, we can fit the marginal models and perform our transformation to standard Laplace margins.
The `cecl_marg` function provides several in-built options for marginal fitting, specified using the `marg_method` argument, including:
- Empirical CDF (`"ecdf"`), which uses the empirical CDF (ECDF) for the entire distribution,
- stationary GPD (`"ismev"`), which fits a GPD to the upper tail of each variable at each location using the `ismev` package [-@ismev],
- non-stationary GPD (`"evgam"`), which fits a GPD to the upper tail of each variable at each location using the `evgam` package [-@evgam], allowing for covariate effects through generalized additive models (GAMs).
Here, we demonstrate how to use each of these marginal fitting methods with the `cecl_marg` function (with the `quantile` thresholding method):
```{r, marginal_fit}
# 1: ECDF
marg_ecdf <- cecl_marg(
  data,
  thresh_method = "quantile",
  thresh_args = 0.9,
  marg_method = "ecdf",
  ncores = 1
)

# 2: ismev
marg_ismev <- cecl_marg(
  data,
  thresh_method = "quantile",
  thresh_args = 0.9,
  marg_method = "ismev",
  ncores = 1,
  # ret_obj = FALSE
  ret_obj = TRUE
)

# 3: evgam
marg_evgam <- cecl_marg(
  data,
  thresh_method = "quantile",
  thresh_args = 0.9,
  marg_method = "evgam",
  # model formulae for GPD scale and shape parameters
  marg_args = list(f = list("excess ~ name", "~ name")),
  ncores = 1,
  ret_obj = TRUE
)


```

Several methods are available for the `cecl_marg` class, including `print` and `summary`:
```{r, marginal_methods}
marg_ismev
summary(marg_ismev, n = 5) # show GPD parameters for the first 5 locations
```

Since `summary` shows the fitted GPD parameters for each location, it will return an error for the ECDF method, since no GPD is fitted in that case.

We also have access to all of the usual GPD diagnostic plots, which we plot here for `loc_1` and variable `X1`:
```{r marginal_plots, fig.width=10, fig.height=8}
par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
plot(marg_ismev, which = "pp", loc = "loc_1", var = "X1")
plot(marg_ismev, which = "qq", loc = "loc_1", var = "X1")
plot(marg_ismev, which = "return", loc = "loc_1", var = "X1")
plot(marg_ismev, which = "hist", loc = "loc_1", var = "X1")
```

These plots are also available through `ggplot`:
```{r, marginal_ggplot}
(ggplot(marg_ismev, which = "pp", loc = "loc_1", var = "X1") + 
   ggplot(marg_ismev, which = "qq", loc = "loc_1", var = "X1")) / 
  (ggplot(marg_ismev, which = "return", loc = "loc_1", var = "X1") + 
     ggplot(marg_ismev, which = "hist", loc = "loc_1", var = "X1"))
```

Unsurprisingly, since we specified GP margins when simulating the data, the GPD fits appear to be very good, even with just 10 observations in the tail (10% of 1000 observations per location).
Future versions of the package will also include methods for visualising the standard Laplace transformed data.

However, We also recognise that marginal fitting in extremes can be a complex task, and users may wish to use their own preferred methods for marginal fitting.
Therefore, the `CeCl` package also provides the `to_cecl_marg` function to allow users to provide their own marginally transformed data in standard Laplace margins.
To illustrate the use of this function, we can provide the transformed data from `ismev` and have `to_cecl_marg` return an object of class `cecl_marg`. 
```{r, to_cecl_marg}
data_trans <- marg_ismev$transformed # list, but can be data.frame/tibble
marg_custom <- as_cecl_marg(data_trans)
marg_custom
```

User-specified marginal transformations work the same as those fitted using `cecl_marg` using the ECDF method, in that no GPD parameters are stored, so plotting methods are not available.
Future versions of the package may also allow users to provide their own fitted GPD parameters for each location and variable, with `as_cecl_marg` performing the necessary transformations.


## Dependence Fits

Next, we fit the CE dependence model to each location using the `cecl_dep` function.
```{r, dependence_fit}
dep <- cecl_dep(obj = marg_ismev, cond_prob = 0.9)
```

Simple as that!
Choosing the conditioning probability is an important step in fitting the CE model, as it determines the threshold above which the model is fitted.
In [@otoole2025], following from the `texmexMultivariate` vignette for [@texmex], bootstrapped estimates of $\alpha$ and $\beta$ across a range of conditioning quantiles were used to identify a suitable conditioning level, as the highest quantile at which the estimates stabilised.
Future versions of the package will include a bootstrapping procedure to assist with this task.

The `cecl_dep` function contains several arguments to control the fitting procedure, including:
- `start`: a list of starting values for the CE parameters $\alpha$ and $\beta$. In the future, we will allow different starting values for each location.
- `nruns`: the number of optimisation runs to perform for each location.
- `aLow`: lower bound for $\alpha$ parameters. For instance, we may wish to restrict $\alpha$ to be non-negative as we expect positive tail dependence. 
- `fixed_b`: a logical indicating whether to fix $\beta$ parameters to it's starting value, which can be useful for interpretation by estimating only the $\alpha$ parameters.
- `fit_no_keef`: a logical indicating whether to fit the model without the Keef et al. (2013) constraints [-@Keef2013], which can sometimes be too restrictive and cause optimisation issues.
Furthermore, for Linux and Mac users with multiple cores available, the fitting can be parallelised using the `ncores` argument. This is also true of all functions including an `ncores` argument, including `cecl_marg`. 

The `cecl_dep` function returns an object of class `cecl_dep`, which has several methods available:
```{r, dependence_methods}
dep
head(summary(dep))
```

Future versions of the package will also include a `predict` method for generating conditional simulations from the fitted CE models, as found in the `texmex` package [@texmex].

We also have access to several diagnostic plots for the fitted CE models.
`residual` plots the residuals $Z$ against large values of the conditioning variable,
`quantile` plots quantiles of the fitted CE model against the conditioning variable (on the standard Laplace scale),
`scatter` plots the $alpha$ and $beta$ parameters against each other for a given variable and conditioning variable.
```{r}
par(mfrow = c(1, 3))
plot(dep, which = "residual", var = "X1", cond_var = "X2", loc = "loc_1")
plot(dep, which = "quantile", var = "X1", cond_var = "X2", loc = "loc_1")
plot(dep, which = "scatter", var = "X1", cond_var = "X2")
```

The residual plot should roughly show no obvious patterns (as indicated by the horizontal trend), and the quantile plot should look sensible and have no lines crossing each other.
Again, these plots are also available through `ggplot`:
```{r, dependence_ggplot}
ggplot(dep, which = "residual", var = "X1", cond_var = "X2", loc = "loc_1") + 
  ggplot(dep, which = "quantile", var = "X1", cond_var = "X2", loc = "loc_1") +
  ggplot(dep, which = "scatter", var = "X1", cond_var = "X2")
```

It is a good idea to print these diagnostic plots for all locations and variable combinations to ensure that the fits are sensible before proceeding to clustering.
For example, to print all residual and quantile plots to a PDF, you could run something like:
```r
pdf("cecl_dep_diagnostics.pdf", width = 8, height = 6)
for (loc in unique(data$name)) {
  ggplot(dep, which = "residual", var = "X1", cond_var = "X2", loc = loc) + 
  ggplot(dep, which = "quantile", var = "X1", cond_var = "X2", loc = loc)
}
dev.off()
```

## Clustering

Finally, we can cluster the fitted CE models using the skew-geometric Jensen-Shannon based dissimilarity measure proposed in [@otoole2025].
First, to calculate the dissimilarity matrix, we can use the `cecl_dist` function:
```{r, distance_matrix}
dist <- cecl_dist(dep, marg_ismev, seed = 123)
```

`cecl_dist` also has several arguments to control the distance calculation, including:
- `laplace_cap`: The divergence is estimating using Monte Carlo integration, simulating values of the conditioning variable from a standard Laplace distribution.
To avoid issues with extremely large values on the standard Laplace scale, we cap the maximum quantile used for simulation at `laplace_cap` (default 0.99).
- `n_mc`: The number of Monte Carlo samples to use for estimating the divergence.
- `par_dist`: An optional argument to parallelise the distance calculation itself, rather than parallelising across conditioning variables. Useful when the number of locations is large.

The `cecl_dist` function returns an object of class `cecl_dist`, which has several methods available:
```{r}
dist
summary(dist)
```

We can produce a scree plot of the total within-group sum of squares (TWGSS) for different numbers of clusters `k`, as well as a heatmap/image of the dissimilarity matrix:
```{r}
par(mfrow = c(1, 2))
invisible(plot(dist, which = "scree")) # normally prints TWGSS for each k
plot(dist, which = "image")
```

And also in `ggplot`:
```{r}
ggplot(dist, which = "scree")$plot + 
  ggplot(dist, which = "image")
```

Unsurprisingly, the scree plot should show a clear "elbow" at the true number of clusters (3 in this case), and the dissimilarity matrix should show three distinct blocks along the diagonal.

Finally, we can perform clustering using the `cecl_clust` function:
```{r, clustering}
clust <- cecl_clust(
  dist, 
  k = 3, 
  cluster_mem = sort(rep(1:3, each = 4)), # known cluster membership
  seed = 123
)
```

Optionally, we can also run `cecl_clust` on the `cecl_dep` and `cecl_marg` objects directly, without needing to calculate the distance matrix first:
```{r}
clust_dep <- cecl_clust(
  dep,
  marg_ismev,
  k = 3,
  cluster_mem = sort(rep(1:3, each = 4)), # known cluster membership
  seed = 123
)

all.equal(unclass(clust), unclass(clust_dep[names(clust_dep) != "dist_obj"]))
```

`clust_dep$dist_obj` will be identical to the output of `cecl_dist` calculated previously, in case the user wishes to inspect it.

The `cecl_clust` function returns an object of class `cecl_clust`, which has several methods available:
```{r}
clust
summary(clust)
```

We can see that the clustering has perfectly recovered the true clusters based on the Adjusted Rand Index (ARI) of 1.
The `summary` method is that of `pam`. 
However, since silhouette widths are not applicable for our dissimilarity measure, please ignore this part of the output (we may remove this in future versions).

We can also produce the `scatter` and `image` plots from previous sections, but this time with the locations ordered and coloured by their assigned cluster:
```{r}
par(mfrow = c(1, 2))
plot(clust, which = "scatter", var = "X1", cond_var = "X2")
plot(clust, which = "image")
```

And also in `ggplot`:
```{r, ggplot_clust}
ggplot(clust, which = "scatter", var = "X1", cond_var = "X2") + 
  (ggplot(clust, which = "image") + 
     theme(
      axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 10),
      legend.title = element_text(size = 12)
     ))
```

In conclusion, we have successfully used the `CeCl` package to cluster multivariate data based on their tail dependence structures using the conditional extremes model.
Please refer to the package documentation for more details on the available functions and methods, 
or contact the author for any questions or suggestions.

## References